---
title: "RProgrammingAssignment3"
author: "Marina W. Kim"
date: "May 26, 2016"
output: html_document
---

The data for this assignment come from the Hospital Compare web site (http://hospitalcompare.hhs.gov) run by the U.S. Department of Health and Human Services. The purpose of the web site is to provide data and information about the quality of care at over 4,000 Medicare-certified hospitals in the U.S. This dataset es- sentially covers all major U.S. hospitals. This dataset is used for a variety of purposes, including determining whether hospitals should be fined for not providing high quality care to patients (see http://goo.gl/jAXFX for some background on this particular topic).

The Hospital Compare web site contains a lot of data and we will only look at a small subset for this assignment. The zip file for this assignment contains three files
*outcome-of-care-measures.csv: Contains information about 30-day mortality and readmission rates for heart attacks, heart failure, and pneumonia for over 4,000 hospitals.
*hospital-data.csv: Contains information about each hospital.
*Hospital_Revised_Flatfiles.pdf: Descriptions of the variables in each file (i.e the code book).

## 1. Plot the 30-day mortality rates for heart attack
```{r}
setwd("~/datasciencecoursera/ProgrammingAssignment3") # setting working directory
outcome <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
# head(outcome) ## looking at the first few rows of data (in comment form to reduce output)
ncol(outcome) # number of columns
nrow(outcome) # number of rows
# names(outcome) ## variable names (in comment form to reduce output)
```

Simple histogram of the 30-day death rates from heart attack:
```{r}
outcome[, 11] <- as.numeric(outcome[, 11])
hist(outcome[, 11], ylab = "Frequency", xlab = "30-day Death Rates", 
     main = "US Hospitals: 30-day Death Rates from Heart Attack")
```


## 2. Finding the best hospital in a state
Writing a function "best" to produce output for the hospital that ranks the highest for a specified outcome measure (heart attack, heart failure, pneumonia) in a state.

Notes about the function:
* If the state or outcome value passed to function "best" is invalid, the function will return an error message and stop evaluating the function.
* If there is a tie for the best hospital for a given outcome, the first hospital name in alphabetical order will be returned. A number will be printed to indicate the number of hospitals in this tie.

```{r}
best <- function(state, outcome){
    outcome_data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
    
    # checking if state value is valid
    matched <- match(state, data[,2], nomatch = 0)
    if (matched == 0){
        stop("invalid state")
    }
    
    # subsetting dataset based on desired outcome analysis
    state_data <- outcome_data[outcome_data$State == state,]
    if(outcome == "heart attack") {
        rank_data0 <- state_data[, c(2, 7, 11)]
    } else if(outcome == "heart failure"){
        rank_data0 <- state_data[, c(2, 7, 17)]
    } else if(outcome == "pneumonia"){
        rank_data0 <- state_data[, c(2, 7, 23)]
    } else {
        stop("invalid outcome")
    }
    
    rank_data <- rank_data0[rank_data0[,3] != "Not Available",]
    rank_data[,3] <- sapply(rank_data[,3], as.numeric)
    
    # determining the lowest outcome rate within a state
    lowest_rate <- min(rank_data[,3], na.rm = TRUE)
    
    # subsetting hospitals with the lowest outcome rate
    competing_hospitals <- rank_data[rank_data[,3] == lowest_rate & !is.na(rank_data[,3]), ]
    
    # sorting alphabetically by hospital name
    competing_hospitals2 <- competing_hospitals[,1]
    competing_hospitals3 <- sort(competing_hospitals2)
    
    if (length(competing_hospitals3) > 1){
        length(competing_hospitals3)
        competing_hospitals3[1] 
    } else {
        competing_hospitals3[1] 
    }
}
```

## 3. Ranking hospitals by outcome in a state

Writing a function "rankhospital" to produce the name of a hospital based on a specified state, outcome, and ranking number.

```{r}
rankhospital <- function(state, outcome, num = "best"){
    outcome_data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
    
    # checking if state value is valid
    matched <- match(state, data[,2], nomatch = 0)
    if (matched == 0){
        stop("invalid state")
    }
    
    # subsetting dataset based on desired outcome analysis
    state_data <- outcome_data[outcome_data$State == state,]
    if(outcome == "heart attack") {
        rank_data0 <- state_data[, c(2, 7, 11)]
    } else if(outcome == "heart failure"){
        rank_data0 <- state_data[, c(2, 7, 17)]
    } else if(outcome == "pneumonia"){
        rank_data0 <- state_data[, c(2, 7, 23)]
    } else {
        stop("invalid outcome")
    }
    
    
    rank_data <- rank_data0[rank_data0[,3] != "Not Available",]
    rank_data[,3] <- sapply(rank_data[,3], as.numeric)
    rank_data2 <- rank_data[order(rank_data[,3], rank_data[,1], 
                                  decreasing = FALSE, na.last = TRUE),]
    rank_data3 <- rank_data2[rank_data2[,3] != "Not Available", ]
    
    # printing the num-ranked hospital
    if (num == "best") {
        rank_data3[1,1]
    } else if (num == "worst"){
        rank_data3[nrow(rank_data3), 1]
    } else {
        rank_data3[num, 1]
    }
}
```

## 4. Ranking hospitals in all states

Writing a function for producing the hospital name and state for all US states, given a specified outcome and ranking number (within the state).

```{r}
rankall <- function(outcome, num = "best"){
    # loading dataset
    outcome_data <- read.csv("outcome-of-care-measures.csv", colClasses = "character")
    
    # subsetting dataset based on desired outcome analysis
    if(outcome == "heart attack") {
        rank_data0 <- outcome_data[, c(2, 7, 11)]
    } else if(outcome == "heart failure"){
        rank_data0 <- outcome_data[, c(2, 7, 17)]
    } else if(outcome == "pneumonia"){
        rank_data0 <- outcome_data[, c(2, 7, 23)]
    } else {
        stop("invalid outcome")
    }
    
    rank_data <- rank_data0[rank_data0[,3] != "Not Available",]
    rank_data[,3] <- sapply(rank_data[,3], as.numeric)
    
    states <- sort(rank_data[,2][!duplicated(rank_data[,2])])
    
    master_rank <- data.frame("Hospital Name" = character(), "State" = character(),
                              "Outcome Rate" = numeric(), "Rank" = numeric())
    # ranking hospitals in each state
    for (i in 1:length(states)){
        state_data <- rank_data[rank_data$State == states[i],]
        complete_state <- state_data[state_data[,3] != "Not Available",]
        ranked_state <- complete_state[order(complete_state[,3], complete_state[,1],
                                                            decreasing = FALSE), ]
        rankings <- c(1:nrow(ranked_state))
        state_rank <- cbind(ranked_state, rankings)
        master_rank <- rbind(master_rank, state_rank)
    }
    
     # printing the num-ranked hospital
    if (num == "best") {
        best <- master_rank[master_rank$rankings == 1, ]
        best_final <- best[, c(1,2)]; best_final
    } else if (num == "worst"){
        worst <- aggregate(rankings ~ State, master_rank, max)
        merged <- merge(master_rank, worst)
        merged2 <- merged[, c(3, 1, 4, 2)]
        worst_final <- merged2[, c(1,2)]; worst_final
    } else {
        specified <- master_rank[master_rank$rankings == num, ]
        specified_final <- specified[, c(1,2)]; specified_final
    }
}
```
